---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
version <- as.character(packageVersion("turnoveR"))
```

# turnoveR

[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/turnoveR)](https://cran.r-project.org/package=turnoveR)
[![Git_Hub_Version](https://img.shields.io/badge/GitHub-`r version`-orange.svg?style=flat-square)](/commits)
[![Last-Update](https://img.shields.io/badge/updated-`r gsub('-', '--', Sys.Date())`-yellowgreen.svg)](/commits)
[![Documentation](https://img.shields.io/badge/docs-online-green.svg)](https://kopflab.github.io/turnoveR/)
[![Build Status](https://travis-ci.org/KopfLab/turnoveR.svg?branch=master)](https://travis-ci.org/KopfLab/turnoveR)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/KopfLab/turnoveR?branch=master&svg=true)](https://ci.appveyor.com/project/KopfLab/turnoveR)

## About

The [turnoveR](https://kopflab.github.io/turnoveR/) package is a collection of tools for the analysis of protein degradation data sets.

## Installation

You can install [turnoveR](https://kopflab.github.io/turnoveR/) from github with the devtools package. If you do not already have many of the dependencies installed, this may take a moment.

```{r gh-installation, eval = FALSE}
# install.packages("devtools") 
devtools::install_github("KopfLab/turnoveR")
```

## Functionality

For an example of how to use the turnoveR package, see the vignette on [Data Analysis](https://kopflab.github.io/turnoveR/articles/analysis_demo.html).
