---
title: "turnoveR: Package Structure"
output: 
  html_document:
    toc: false
pkgdown:
  as_is: true
---

```{r echo=FALSE}
# enable font awesome
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

The following flow chart illustrates the main functions and work flow of the turnoveR package. Mouseover individual steps to get additional information and click on individual turnover (`tor_`) functions to jump directly to the function documentation. This flow chart can also be easily generated from within RStudio as a on-the-fly reference using the `tor_show_package_structure()` function.

## Legend

 - <i class="fa fa-file-o"/> data files colored in <font color="blue">**blue**</font> - <i class="fa fa-file-text-o"/> text/csv files, <i class="fa fa-file-excel-o"/> Excel files, and <i class="fa fa-file-archive-o"/> complex data files
 - <i class="fa fa-bolt"/> functions colored in <font color="green">**green**</font> if already implemented, in <font color="red">**red**</font> if planned but not yet implemented
 - <i class="fa fa-calculator"/> stand-alone external programs in <font color="orange">**orange**</font>
 - <i class="fa fa-database"/> online databases
 - <i class="fa fa-info"/> individual additional information from other sources 
 
## Flow Chart

```{r}
turnoveR::tor_show_package_structure(height = 1100)
```

